Global optimization: from one-dimensional to multidimensional problems via the dimensionality reduction

### 2017-10-06 2017/ParOptBook/Chapter1.tex
Chapter 6 is ready
Some misprints have been corrected
Chapters 3 and 5 have been included. There is a question about cross-references between chapters.

\abstract{This chapter is devoted to the basic statements and general definitions concerning the global optimization problems and algorithms. At the beginning, relevant notions are introduced and different statements of global optimization problems are described. As a main optimization statement, a non-linear programming problem is considered.  The concepts of numerical methods for solving the optimization problems are discussed and a general abstract model for successive methods is introduced. On the base of abstract description its algorithmic implementation (general computational scheme) is given. The model of sequential optimization expands for the case of parallel computations and general computational procedures for synchronous and asynchronous parallel methods are presented. Convergence of optimization methods and possibility of obtaining estimations of the problem solution after a finite number of objective function evaluations are discussed. Significant complexity of multidimensional multie
xtremal optimization problems is considered and approaches of overcoming this complexity via reduction schemes are briefly described. }

\section{The global optimization problem}
\label{sec:1}
The aim of this section is to introduce the necessary terms and to formulate the general problem of the global optimization, the investigation of which the present monograph is dedicated to. 
Let $\varphi(y)$ be a real-valued function, defined over a domain $D$ of the $N$-dimensional Euclidean space $R^N$ with a finite value at every point of $Q$.

@@

Let us consider an optimization problem  for the function
\begin{displaymath}
\label{eq:12}
\varphi(y_1,y_2)=-1.5y_1^2\exp(1-y_1^2-20.25(y_1-y_2)^2)-
\end{displaymath}
\begin{displaymath}
\label{eq:12}
-(0.5(y_1-1)(y_2-1))^4\exp(2-(0.5(y_1-1))^4-(y_2-1)^4)
\end{displaymath}
within the domain $0 \leq y_1\leq 4,\ \  -1\leq y_2\leq 3$, with the following constraints 
\begin{displaymath}
g_1(y_1,y_2)=0.01((y_1-2.2)^2+(y_2-1.2)^2-2.25)\leq 0,
\end{displaymath}
\begin{displaymath}
g_2(y_1,y_2)=100(1-(y_1-2)^2/1.44-(0.5y_2)^2)\leq 0,
\end{displaymath}
\begin{displaymath}
g_3(y_1,y_2)=10(y_2-1.5-1.5\sin(6.283(y_1-1.75)))\leq 0,
\end{displaymath}
as a particular example.

@@

The set of the points satisfying the first constraint represents a circle with the boundary defined by the equation $g_1(y_1,y_2)=0$. The points satisfying the second constraint are located outside an ellipse defined by the equation $g_2(y_1,y_2)=0$. The points satisfying the third constraint are below the sinusoid $g_3(y_1,y_2)=0$. Thus, the feasible domain is disconnect and consists of three non-convex subdomains (the feasible domain is dark in Fig.~\ref{1_fig_1}). The global minimum $\varphi(y_1^*,y_2^*)=-1.489$  is achieved at the point $(y_1^*,y_2^*)$  = (0,942, 0,944).

\section{Numerical optimization methods}
\label{sec:1.2}

@@

For this method
\begin{displaymath}
G_1(\Phi)=a, G_{k=1}(\Phi,\omega_k)=a+k\frac{b-a}{q}, k\geq 1,
\end{displaymath}
\begin{displaymath}
  H_k(\Phi,\omega_k) =
  \begin{cases}
    0, & k > q, \\
    1, & k \leq q,
  \end{cases}
\end{displaymath}
\begin{equation}
\label{eq:18}
e^k=\varphi_k^*,
\end{equation}
where
\begin{equation}
\label{eq:19}
\varphi_k^*=\min_{1\leq i\leq k} \varphi(y^i),
\end{equation}
or
\begin{displaymath}
e^k=(\varphi_k^*,y_k^*),
\end{displaymath}
where 
\begin{displaymath}
\label{eq:19}
y_k^*=\arg \min_{1\leq i\leq k} \varphi(y^i).
\end{displaymath}

@@

\section{Multiextremal problems and reduction of complexity}
\label{sec:3}
In multiextremal problems the dimension affects the difficulty of solving such problems significantly. For example, for the class of the functions satisfying Lipschitz condition so called «the curse of dimensionality» consisting in the exponential growth of the computational costs with increasing dimensionality takes place. Namely, if $p$ evaluations of the objective function are required to achieve the solution accuracy $\epsilon$ in the one-dimensional case, in the $N$-dimensional problem $\alpha p^N$  trials are required to achieve the same accuracy. Here the coefficient $\alpha$ depends on the objective function, on the feasible domain, and on the method used.

@@

There also exist other schemes built on the base of the reduction of a multiextremal problem to solving the local subproblems. The algorithms designed within the framework of this approach have an asymptotic convergence to the global extremum with weak presumptions of the continuity or of some degree of differentiability of the objective function. From the quite strict theoretical point of view, this convergence is provided by such property that any point of the search domain is an accumulation point of the trial sequence   generated by the algorithm and, consequently, for a continuous function
\begin{equation}
\label{eq:31}
\lim_{\tau \to \infty}\min_{1\leq k \leq \tau}=\varphi^*.
\end{equation}
In other words, the trial sequences of these methods are everywhere dense in the search domain, i.e., converge in the sense of Definition 1.4 to all points within this domain and, consequently, to the global minimum point as well.

@@

\bibitem{BetroSchoen} Betro, B., Schoen, F.: Sequential stopping rules for the multistart algorithm in global optimisation. Math. Program. \textbf{38}(3), 271-286 (1987)
\bibitem{BoendeRinnooy}	Boender, C.G.E, Rinnooy Kan, A.H.G.: Bayesian stopping rules for multistart global optimization methods. Math. Program. \textbf{37}(1), 59-80 (1987)
\bibitem{1_Butz}	Butz, A.R. Space-Filling Curves and Mathematical Programming.: Inform. Control \textbf{12}, 314-330 (1968)
\bibitem{1_CarrHowe}	Carr, C.R., Howe, C.W. Quantitative Decision Procedures in Management and Economic: Deterministic Theory and Applications. McGraw-Hill, New York (1964)

@@

\bibitem{1_GerGriIsr} Gergel, V.,  Grishagin, V., Israfilov, R.: Local tuning in nested scheme of global optimization, Procedia Computer Science \textbf{51}, 865-874 (2015) 

@@

\bibitem{1_Goertzel}	Goertzel, B.: Global Optimization with Space-Filling Curves. Appl. Math. Lett. 12, 133-135 (1999)

@@

\bibitem{1_GriIsrCEUR}Grishagin, V.A., Israfilov, R.A. Multidimensional Constrained Global Optimization in Domains with Computable Boundaries. CEUR Workshop Proceedings 1513, 75-84 (2015)
\bibitem{1_HimOliPet}	Hime, A.E., Oliveira Jr., H.A., Petraglia, A.: Global Optimization Using Space-Filling Curves and Measure-Preserving Transformations. Soft Computing in Industrial Applications \textbf{96}, 121-130 (2011)

@@

\bibitem{1_Kushner}	Kushner, H.J.: A new method of locating the maximum point of an arbitrary multipeak curve in the presence of noise. Transactions of ASME, Ser. D. Journal of Basic Engineering \textbf{86}, 97-106 (1964)
\bibitem{1_LeraSergCNSNS}	Lera, D., Sergeyev, Y.D.: Deterministic global optimization using space-filling curves and multiple estimates of Lipschitz and H{\"o}lder constants. Communications in Nonlinear Science and Numerical Simulation \textbf{23}, 328-342 (2015)
\bibitem{1_LeraSergANM}	Lera, D., Sergeyev, Y.D.: Lipschitz and H{\"o}lder global optimization using space-filling curves. Applied Numerical Mathematics \textbf{60}, 115-129 (2010) 

@@

\bibitem{1_Piyavskij}	Piyavskij, S.A.: An Algorithm for Finding the Absolute Extremum of a Function. USSR Comput. Math. Math. Phys. \textbf{12}(4), 57-67 (1972)
\bibitem{1_SergGriJCAA}	Sergeyev, Y.D., Grishagin, V.A.: Parallel Asynchronous Global Search and the Nested Optimization Scheme. J. Comp. Analysis Appl. \textbf{3}, 123-145 (2001)

@@

\bibitem{1_ShiOlaf}   Shi, L., {\'O}lafsson, S.: Nested Partitions Method for Global Optimization. Operations Research \textbf{48}, 390-407 (2000)

@@

\bibitem{1_vanDam}	van Dam, E.R., Husslage, B., Hertog, D. One-dimensional Nested Maximin Designs. J. Glob. Opt. \textbf{46}, 287-306 (2010)

@@

\bibitem{1_ZhigZhil} Zhigljavsky, A.A., Zilinskas, A.: Stochastic Global Optimization. Springer, New York. (2008)
\bibitem{1_Zielinski}	Zieli{\'n}ski, R.: A statistical estimate of the structure of multi-extremal problems. Math. Program. \textbf{21}(1), 348-356 (1981)

>>>>

\abstract{This chapter is devoted to the basic statements and general definitions concerning the global optimization problems and algorithms. At the beginning, relevant notions are introduced and different statements of global optimization problems are described. As a main optimization statement, a non-linear programming problem is considered.  The concepts of numerical methods for solving the optimization problems are discussed and a general abstract model for successive methods is introduced. On the base of abstract description its algorithmic implementation (general computational scheme) is given. The model of sequential optimization expands for the case of parallel computations and general computational procedures for synchronous and asynchronous parallel methods are presented. Convergence of optimization methods and possibility of obtaining estimations of the problem solution after a finite number of objective function evaluations are discussed. Significant complexity of multidimensional multiextremal optimization problems is considered and approaches of overcoming this complexity via reduction schemes are briefly described. }

\section{The global optimization problem}
\label{sec:1_1}
The aim of this section is to introduce the necessary terms and to formulate the general problem of the global optimization, the investigation of which the present monograph is dedicated to. 
Let $\varphi(y)$ be a real-valued function, defined over a domain $D$ of the $N$-dimensional Euclidean space $R^N$ with a finite value at every point of $Q$.

@@

Let us consider an optimization problem  for the function
\begin{displaymath}
%\label{eq:12}
\varphi(y_1,y_2)=-1.5y_1^2\exp(1-y_1^2-20.25(y_1-y_2)^2)-
\end{displaymath}
\begin{displaymath}
%\label{eq:12}
-(0.5(y_1-1)(y_2-1))^4\exp(2-(0.5(y_1-1))^4-(y_2-1)^4)
\end{displaymath}
within the domain $0 \leq y_1\leq 4,\ \  -1\leq y_2\leq 3$, with the following constraints 
\begin{displaymath}
g_1(y_1,y_2)=0.01((y_1-2.2)^2+(y_2-1.2)^2-2.25)\leq 0,
\end{displaymath}
\begin{displaymath}
g_2(y_1,y_2)=100(1-(y_1-2)^2/1.44-(0.5y_2)^2)\leq 0,
\end{displaymath}
\begin{displaymath}
g_3(y_1,y_2)=10(y_2-1.5-1.5\sin(6.283(y_1-1.75)))\leq 0,
\end{displaymath}
as a particular example.

@@

The set of the points satisfying the first constraint represents a circle with the boundary defined by the equation $g_1(y_1,y_2)=0$. The points satisfying the second constraint are located outside an ellipse defined by the equation $g_2(y_1,y_2)=0$. The points satisfying the third constraint are below the sinusoid $g_3(y_1,y_2)=0$. Thus, the feasible domain is disconnect and consists of three non-convex subdomains (the feasible domain is dark in Fig.~\ref{1_fig_1}). The global minimum $\varphi(y_1^*,y_2^*)=-1.489$  is achieved at the point $(y_1^*,y_2^*)$  = (0.942, 0.944).

\section{Numerical optimization methods}
\label{sec:1_2}

@@

For this method
\begin{displaymath}
G_1(\Phi)=a, G_{k=1}(\Phi,\omega_k)=a+k\frac{b-a}{q}, k\geq 1,
\end{displaymath}
\begin{displaymath}
  H_k(\Phi,\omega_k) =
  \begin{cases}
    0, & k > q, \\
    1, & k \leq q,
  \end{cases}
\end{displaymath}
\begin{equation}
\label{eq:18}
e^k=\varphi_k^*,
\end{equation}
where
\begin{equation}
\label{eq:19}
\varphi_k^*=\min_{1\leq i\leq k} \varphi(y^i),
\end{equation}
or
\begin{displaymath}
e^k=(\varphi_k^*,y_k^*),
\end{displaymath}
where 
\begin{displaymath}
%\label{eq:19}
y_k^*=\arg \min_{1\leq i\leq k} \varphi(y^i).
\end{displaymath}

@@

\section{Multiextremal problems and reduction of complexity}
\label{sec:1_3}
In multiextremal problems the dimension affects the difficulty of solving such problems significantly. For example, for the class of the functions satisfying Lipschitz condition so called «the curse of dimensionality» consisting in the exponential growth of the computational costs with increasing dimensionality takes place. Namely, if $p$ evaluations of the objective function are required to achieve the solution accuracy $\epsilon$ in the one-dimensional case, in the $N$-dimensional problem $\alpha p^N$  trials are required to achieve the same accuracy. Here the coefficient $\alpha$ depends on the objective function, on the feasible domain, and on the method used.

@@

There also exist other schemes built on the base of the reduction of a multiextremal problem to solving the local subproblems. The algorithms designed within the framework of this approach have an asymptotic convergence to the global extremum with weak presumptions of the continuity or of some degree of differentiability of the objective function. From the quite strict theoretical point of view, this convergence is provided by such property that any point of the search domain is an accumulation point of the trial sequence   generated by the algorithm and, consequently, for a continuous function
\begin{equation}
\label{eq:31}
\lim_{\tau \to \infty}\min_{1\leq k \leq \tau}{\varphi(y^k)}=\varphi^*.
\end{equation}
In other words, the trial sequences of these methods are everywhere dense in the search domain, i.e., converge in the sense of Definition \ref{def:1_4} to all points within this domain and, consequently, to the global minimum point as well.

@@

\bibitem{BetroSchoen} Betro, B., Schoen, F.: Sequential stopping rules for the multistart algorithm in global optimisation. Math. Program. \textbf{38}(3), 271--286 (1987)
\bibitem{BoendeRinnooy}	Boender, C.G.E, Rinnooy Kan, A.H.G.: Bayesian stopping rules for multistart global optimization methods. Math. Program. \textbf{37}(1), 59--80 (1987)
\bibitem{1_Butz}	Butz, A.R. Space-Filling Curves and Mathematical Programming.: Inform. Control \textbf{12}, 314--330 (1968)
\bibitem{1_CarrHowe}	Carr, C.R., Howe, C.W.: Quantitative Decision Procedures in Management and Economic: Deterministic Theory and Applications. McGraw-Hill, New York (1964)

@@

\bibitem{1_GerGriIsr} Gergel, V.,  Grishagin, V., Israfilov, R.: Local tuning in nested scheme of global optimization, Procedia Computer Science \textbf{51}, 865--874 (2015) 

@@

\bibitem{1_Goertzel}	Goertzel, B.: Global Optimization with Space-Filling Curves. Appl. Math. Lett. 12, 133--135 (1999)

@@

\bibitem{1_GriIsrCEUR}Grishagin, V.A., Israfilov, R.A.: Multidimensional Constrained Global Optimization in Domains with Computable Boundaries. CEUR Workshop Proceedings 1513, 75--84 (2015)
\bibitem{1_HimOliPet}	Hime, A.E., Oliveira Jr., H.A., Petraglia, A.: Global Optimization Using Space-Filling Curves and Measure-Preserving Transformations. Soft Computing in Industrial Applications \textbf{96}, 121--130 (2011)

@@

\bibitem{1_Kushner}	Kushner, H.J.: A new method of locating the maximum point of an arbitrary multipeak curve in the presence of noise. Transactions of ASME, Ser. D. Journal of Basic Engineering \textbf{86}, 97--106 (1964)
\bibitem{1_LeraSergCNSNS}	Lera, D., Sergeyev, Y.D.: Deterministic global optimization using space-filling curves and multiple estimates of Lipschitz and H{\"o}lder constants. Communications in Nonlinear Science and Numerical Simulation \textbf{23}, 328--342 (2015)
\bibitem{1_LeraSergANM}	Lera, D., Sergeyev, Y.D.: Lipschitz and H{\"o}lder global optimization using space-filling curves. Applied Numerical Mathematics \textbf{60}, 115--129 (2010) 

@@

\bibitem{1_Piyavskij}	Piyavskij, S.A.: An Algorithm for Finding the Absolute Extremum of a Function. USSR Comput. Math. Math. Phys. \textbf{12}(4), 57--67 (1972)
\bibitem{1_SergGriJCAA}	Sergeyev, Y.D., Grishagin, V.A.: Parallel Asynchronous Global Search and the Nested Optimization Scheme. J. Comp. Analysis Appl. \textbf{3}, 123--145 (2001)

@@

\bibitem{1_ShiOlaf}   Shi, L., {\'O}lafsson, S.: Nested Partitions Method for Global Optimization. Operations Research \textbf{48}, 390--407 (2000)

@@

\bibitem{1_vanDam}	van Dam, E.R., Husslage, B., Hertog, D. One-dimensional Nested Maximin Designs. J. Glob. Opt. \textbf{46}, 287--306 (2010)

@@

\bibitem{1_ZhigZhil} Zhigljavsky, A.A., $\check{Z}$ilinskas, A.: Stochastic Global Optimization. Springer, New York. (2008)
\bibitem{1_Zielinski}	Zieli{\'n}ski, R.: A statistical estimate of the structure of multi-extremal problems. Math. Program. \textbf{21}(1), 348--356 (1981)
